Fitness traits of deoxynivalenol and nivalenol-producing Fusarium graminearum species complex strains from wheat

### 2017-12-18 manuscript.Rmd
Several changes including .doc as output

---
title: "Fitness traits of deoxynivalenol and nivalenol-producing Fusarium graminearum species complex strains from wheat"
bibliography: crossref.bib
csl: plant-disease.csl
---

Camila P. Nicolli, Franklin J. Machado, Piérri Spolti, Emerson M. Del Ponte   
Departamento de Fitopatologia, Universidade Federal de Viçosa, 36570-000, Viçosa, MG, Brazil.

@@

> *Fusarium graminearum* of the 15-acetyl(A)deoxynivalenol(DON) chemotype is the main cause of Fusarium head blight (FHB) of wheat in south of Brazil. However, 3-ADON and nivalenol(NIV) chemotypes have been found in other members of the species complex causing FHB in wheat. To improve our understanding of the pathogen ecology, we assessed a range of fitness-related traits in a sample of 30 strains representatives of 15-ADON (**F. graminearum**), 3-ADON (*F. cortaderiae* and *F. austroamericanum*) and NIV (*F. meridionale* and *F. cortaderiae*). These included: perithecia formation on three cereal-based substrates, mycelial growth at two suboptimal temperatures, sporulation and germination, pathogenicity towards a susceptible and a moderately resistant cultivar and sensitivity to tebuconazole. The most important trait favoring *F. graminearum* was its 2x higher sexual fertility (> 40% PPI = perithecia production index) than the other species (< 30% PPI); PPI varied among substrates (maize > rice > wheat). In addition, sensitivity to tebuconazole appeared lower in *F. graminearum* which had the only strain with EC50 > 1 ppm. In the pathogenicity assays, the DON-producers were generally more aggressive (1.5 to 2x higher final severity) towards the two cultivars, with 3-ADON or 15-ADON leading to higher area under the severity curve than the NIV strains in the susceptible and moderately resistant cv., respectively. There was significant variation among strains of a same species with regards asexual fertility (mycelial growth, macroconidia production and germination), which suggest a strain- rather than a species-specific differences. These results contribute new knowledge to improve our understanding of the pathogen-related traits that may explain the dominance of certain members of the species complex in specific wheat agroecosystems.

Key words: Triticum aestivum, trichothecenes, fitness, FGSC.

@@

Fusarium head blight (FHB) is a major disease of wheat and small grain crops due to the reduction in yield and contamination of grain with dangerous mycotoxins (McMullen et al. 2012). Deoxynivalenol (DON) is the most significant mycotoxin given its widespread occurrence at levels of concern to human and animal health (Tanaka et al. 1988). However, nivalenol (NIV), another type-B trichothecene, has been found in commercial wheat samples from Brazil (Del Ponte et al. 2015) and also in inoculations of wheat under controlled conditions (Mendes et al. 2018; Nicolli et al. 2015). A molecular survey conducted during the last decade showed that *Fusarium graminearum* (O’Donnell et al. 2004), formerly lineage 7 (O’Donnell et al. 2000), is the main (>85% frequency) *F. graminearum* species complex (FGSC) member causing FHB of wheat in Brazil (Del Ponte et al. 2015). The **F. graminearum** strains found in Brazilian wheat and barley analyzed so far are exclusively of the 15-acetyl(A)DON genotype, a mainly DON producer (Astolfi et al. 2012; Castañares et al. 2016; Del Ponte et al. 2015). *Fusarium meridionale* is the second most dominant FHB pathogen (>15% overall) and an important regional contributor of FHB in the major wheat-producing region of Brazil, Paraná State, where its frequency increases to ~30% in wheat spikes and grain (Del Ponte et al. 2015). This species is the main cause (> 96% frequency) of Gibberella ear rot and stalk rot as well the most frequent species surviving on maize stubbles, a large reservoir of inoculum for FHB epidemics (Del Ponte et al. 2015; Kuhnem et al. 2015). Other FGSC members that occur at smaller frequency in wheat include *F. cortaderiae* and *F. austroamericanum* that possesses either 3ADON or NIV trichothecene genotype (Del Ponte et al. 2015). Nivalenol, as well as 3-ADON strains within FGSC members are actually much less common in wheat grown in the Americas, especially in Brazil (Del Ponte et al. 2015), but there are concerns about their ability to produce different or greater amount of trichothecenes (Mendes et al. 2018; Umpiérrez-Failache et al. 2013) and potential shifts in climate that may favor build up of inoculum of the currently minor species (Juroszek et al. 2013; Vaughan et al. 2016). In the state of Louisiana, U.S., NIV-type populations of *F. graminearum*, which were found at high frequency (79%) among isolates from small-grain-growing regions, accumulated four times less trichothecene toxin than DON types on inoculated wheat (Gale et al. 2011). The 3-ADON strains of *F. graminearum*, which appeared to emerge and spread into some regions of North America, were able to produce higher quantities of total trichothecenes and exhibit higher fecundity and growth rates than 15-ADON strains (Ward et al. 2008). More recently, the aggressiveness of *F. graminearum* strains of the 15ADON or 3-ADON genotypes from New York towards a susceptible wheat cultivar did not differ, thus suggesting no apparent advantage of 3-ADON compared to 15-ADON genotype, although the former produced significantly greater amount of trichothecenes (Spolti et al. 2014a).

@@

Empirical knowledge of some components of the life cycle of other FGSC members than *F. graminearum* and F. meridionale in Brazil, such as the saprophytic ability, and whether this is influenced by the cereal-based substrate and temperature, which may provide competitive advantages for a species to build up inoculum for further spread are not available. In this study, a sample of 30 strains representative of three trichothecene genotypes among four FGSC members associated with FHB of wheat in southern Brazil was selected to investigate the fitness-related traits related to both the saprophytic and pathogenic phase.

@@

**Table 1.** Information for a collection of 30 strains belonging to four phylogenetic species of the Fusarium graminearum species complex isolated from symptomatic wheat heads from fields in Rio Grande do Sul State, Brazil, which were used in five different experiments in this study

| CML Code<sup>a</sup> | Location             | Year of isolation | Species             | Genotype<sup>b</sup> | Experiment<sup>c</sup> |
|------|----------------------|-------------------|---------------------|----------|------------|
| 3065 | Panambi              | 2009              | F. graminearum      | 15-ADON  | 2, 3, 4    |
| 3066 | Ijuí                 | 2009              | F. graminearum      | 15-ADON  | 1,4        |
| 3067 | Carazinho            | 2010              | F. graminearum      | 15-ADON  | 1,2,3,4    |
| 3071 | Tapejara             | 2010              | F. graminearum      | 15-ADON  | 1,2,3,4,5  |
| 3069 | Coxilha              | 2010              | F. graminearum      | 15-ADON  | 1, 2, 4,5  |
| 3070 | Santa Bárbara do Sul | 2011              | F. graminearum      | 15-ADON  | 1, 2, 4    |
| 3830 | Não-Me-Toque         | 2011              | F. graminearum      | 15-ADON  | 1,2,3      |
| 3833 | Não-Me-Toque         | 2011              | F. graminearum      | 15-ADON  | 1          |
| 3064 | Cruz Alta            | 2007              | F. graminearum      | 15-ADON  | 1,4,5      |
| 3068 | Ernestina            | 2007              | F. graminearum      | 15-ADON  | 1,2,3,4,5  |
| 3382 | Caseiros             | 2010              | F. austroamericanum | 3-ADON   | 1,2,3,5    |
| 3384 | Barreto              | 2010              | F. austroamericanum | 3-ADON   | 1,2,3,4,5  |
| 3379 | Carazinho            | 2010              | F. cortaderiae      | 3-ADON   | 1,2,3,4,5  |
| 3845 | Lagoa Vermelha       | 2010              | F. cortaderiae      | 3-ADON   | 1,2,3,4    |
| 3841 | Lagoa Vermelha       | 2010              | F. cortaderiae      | 3-ADON   | 1,2,3,4,5  |
| 3831 | Panambi              | 2009              | F. cortaderiae      | NIV      | 1,2,3,4,5  |
| 3836 | Coronel Barros       | 2009              | F. cortaderiae      | NIV      | 1,2,4      |
| 3837 | Coronel Barros       | 2009              | F. cortaderiae      | NIV      | 1,2,3,4,5  |
| 3840 | Mato Castelhano      | 2010              | F. cortaderiae      | NIV      | 1,2,4      |
| 3834 | Barreto              | 2010              | F. cortaderiae      | NIV      | 1,2,4      |
| 3378 | Vacaria              | 2010              | F. cortaderiae      | NIV      | 1,2,3,4,5  |
| 3835 | Ernestina            | 2007              | F. cortaderiae      | NIV      | 1,2,3,4    |
| 3375 | Santa Bárbara do Sul | 2009              | F. meridionale      | NIV      | 1,2,4      |
| 3843 | Ernestina            | 2009              | F. meridionale      | NIV      | 1,2,3,4    |
| 3344 | Coxilha              | 2009              | F. meridionale      | NIV      | 1,2,3,4,5  |
| 3381 | Caseiros             | 2010              | F. meridionale      | NIV      | 1,2,3,4,5  |
| 3839 | Marau                | 2011              | F. meridionale      | NIV      | 1,2,3,4,5  |
| 3377 | Sertão               | 2011              | F. meridionale      | NIV      | 1,2,3,4    |
| 3838 | Nonoai               | 2007              | F. meridionale      | NIV      | 1,2,3,4    |
| 3374 | Nonoai               | 2007              | F. meridionale      | NIV      | 1,2,4      |

@@

Spolti, P., Del PontE, E.M. 2013. Agressividade diferencial de espécies do complexo *Fusarium graminearum* em interação com o fungicida tebuconazole na redução do rendimento de trigo. Ciênc. Rural 43:1569-1575.

@@

Spolti, P., Del Ponte, E.M., Dong, Y., Bergstrom, G. 2014b. Triazole sensitivity in a contemporary population of Fusarium graminearum from New York wheat and competitiveness of a tebuconazole-resistant Isolate. Plant Dis. 98:607-613.

@@

Tunali, B., Obanor, F., Erginbas, G., Westecott, R.A., Nicol, J., Chakraborty, S. 2012. Fitness of three Fusarium pathogens of wheat. Microbiol Ecol. 81:596-609.

@@

>>>>

---
title: "Fitness traits of deoxynivalenol and nivalenol-producing Fusarium graminearum species complex strains from wheat"
author: "C. P. Nicolli, F. J. Machado, P. Spolti, E. M. Del Ponte"
csl: plant-disease.csl
bibliography: crossref.bib
output: 
  word_document:
    reference_docx: manuscript.docx
---

Departamento de Fitopatologia, Universidade Federal de Viçosa, 36570-000, Viçosa, MG, Brazil.

@@

> *Fusarium graminearum* of the 15-acetyl(A)deoxynivalenol(DON) chemotype is the main cause of Fusarium head blight (FHB) of wheat in south of Brazil. However, 3-ADON and nivalenol(NIV) chemotypes have been found in other members of the species complex causing FHB in wheat. To improve our understanding of the pathogen ecology, we assessed a range of fitness-related traits in a sample of 30 strains representatives of 15-ADON (*F. graminearum*), 3-ADON (*F. cortaderiae* and *F. austroamericanum*) and NIV (*F. meridionale* and *F. cortaderiae*). These included: perithecia formation on three cereal-based substrates, mycelial growth at two suboptimal temperatures, sporulation and germination, pathogenicity towards a susceptible and a moderately resistant cultivar and sensitivity to tebuconazole. The most important trait favoring *F. graminearum* was its 2x higher sexual fertility (> 40% PPI = perithecia production index) than the other species (< 30% PPI); PPI varied among substrates (maize > rice > wheat). In addition, sensitivity to tebuconazole appeared lower in *F. graminearum* which had the only strain with EC50 > 1 ppm. In the pathogenicity assays, the DON-producers were generally more aggressive (1.5 to 2x higher final severity) towards the two cultivars, with 3-ADON or 15-ADON leading to higher area under the severity curve than the NIV strains in the susceptible and moderately resistant cv., respectively. There was significant variation among strains of a same species with regards asexual fertility (mycelial growth, macroconidia production and germination), which suggest a strain- rather than a species-specific differences. These results contribute new knowledge to improve our understanding of the pathogen-related traits that may explain the dominance of certain members of the species complex in specific wheat agroecosystems.

Key words: *Triticum aestivum*, trichothecenes, fitness, FGSC.

@@

Fusarium head blight (FHB) is a major disease of wheat and small grain crops due to the reduction in yield and contamination of grain with dangerous mycotoxins (McMullen et al. 2012). Deoxynivalenol (DON) is the most significant mycotoxin given its widespread occurrence at levels of concern to human and animal health (Tanaka et al. 1988). However, nivalenol (NIV), another type-B trichothecene, has been found in commercial wheat samples from Brazil (Del Ponte et al. 2015) and also in inoculations of wheat under controlled conditions (Mendes et al. 2018; Nicolli et al. 2015). A molecular survey conducted during the last decade showed that *Fusarium graminearum* (O’Donnell et al. 2004), formerly lineage 7 (O’Donnell et al. 2000), is the main (>85% frequency) *F. graminearum* species complex (FGSC) member causing FHB of wheat in Brazil (Del Ponte et al. 2015). The *F. graminearum* strains found in Brazilian wheat and barley analyzed so far are exclusively of the 15-acetyl(A)DON genotype, a mainly DON producer (Astolfi et al. 2012; Castañares et al. 2016; Del Ponte et al. 2015). *Fusarium meridionale* is the second most dominant FHB pathogen (>15% overall) and an important regional contributor of FHB in the major wheat-producing region of Brazil, Paraná State, where its frequency increases to ~30% in wheat spikes and grain (Del Ponte et al. 2015). This species is the main cause (> 96% frequency) of Gibberella ear rot and stalk rot as well the most frequent species surviving on maize stubbles, a large reservoir of inoculum for FHB epidemics (Del Ponte et al. 2015; Kuhnem et al. 2015). Other FGSC members that occur at smaller frequency in wheat include *F. cortaderiae* and *F. austroamericanum* that possesses either 3ADON or NIV trichothecene genotype (Del Ponte et al. 2015). Nivalenol, as well as 3-ADON strains within FGSC members are actually much less common in wheat grown in the Americas, especially in Brazil (Del Ponte et al. 2015), but there are concerns about their ability to produce different or greater amount of trichothecenes (Mendes et al. 2018; Umpiérrez-Failache et al. 2013) and potential shifts in climate that may favor build up of inoculum of the currently minor species (Juroszek et al. 2013; Vaughan et al. 2016). In the state of Louisiana, U.S., NIV-type populations of *F. graminearum*, which were found at high frequency (79%) among isolates from small-grain-growing regions, accumulated four times less trichothecene toxin than DON types on inoculated wheat (Gale et al. 2011). The 3-ADON strains of *F. graminearum*, which appeared to emerge and spread into some regions of North America, were able to produce higher quantities of total trichothecenes and exhibit higher fecundity and growth rates than 15-ADON strains (Ward et al. 2008). More recently, the aggressiveness of *F. graminearum* strains of the 15ADON or 3-ADON genotypes from New York towards a susceptible wheat cultivar did not differ, thus suggesting no apparent advantage of 3-ADON compared to 15-ADON genotype, although the former produced significantly greater amount of trichothecenes (Spolti et al. 2014a).

@@

Empirical knowledge of some components of the life cycle of other FGSC members than *F. graminearum* and *F. meridionale* in Brazil, such as the saprophytic ability, and whether this is influenced by the cereal-based substrate and temperature, which may provide competitive advantages for a species to build up inoculum for further spread are not available. In this study, a sample of 30 strains representative of three trichothecene genotypes among four FGSC members associated with FHB of wheat in southern Brazil was selected to investigate the fitness-related traits related to both the saprophytic and pathogenic phase.

@@

@@

Spolti, P., Del Ponte, E.M. 2013. Agressividade diferencial de espécies do complexo *Fusarium graminearum* em interação com o fungicida tebuconazole na redução do rendimento de trigo. Ciênc. Rural 43:1569-1575.

@@

Spolti, P., Del Ponte, E.M., Dong, Y., Bergstrom, G. 2014b. Triazole sensitivity in a contemporary population of *Fusarium graminearum* from New York wheat and competitiveness of a tebuconazole-resistant Isolate. Plant Dis. 98:607-613.

@@

Tunali, B., Obanor, F., Erginbas, G., Westecott, R.A., Nicol, J., Chakraborty, S. 2012. Fitness of three *Fusarium* pathogens of wheat. Microbiol Ecol. 81:596-609.

@@

**Table 1.** Information for a collection of 30 strains belonging to four phylogenetic species of the *Fusarium graminearum* species complex isolated from symptomatic wheat heads from fields in Rio Grande do Sul State, Brazil, which were used in five different experiments in this study

| CML <sup>a</sup> | Location             | Year  | Species             | Genotype<sup>b</sup> | Experiment<sup>c</sup> |
|------|----------------------|-------------------|---------------------|----------|------------|
| 3065 | Panambi              | 2009              | F. graminearum      | 15-ADON  | 2, 3, 4    |
| 3066 | Ijuí                 | 2009              | F. graminearum      | 15-ADON  | 1,4        |
| 3067 | Carazinho            | 2010              | F. graminearum      | 15-ADON  | 1,2,3,4    |
| 3071 | Tapejara             | 2010              | F. graminearum      | 15-ADON  | 1,2,3,4,5  |
| 3069 | Coxilha              | 2010              | F. graminearum      | 15-ADON  | 1, 2, 4,5  |
| 3070 | Santa Bárbara do Sul | 2011              | F. graminearum      | 15-ADON  | 1, 2, 4    |
| 3830 | Não-Me-Toque         | 2011              | F. graminearum      | 15-ADON  | 1,2,3      |
| 3833 | Não-Me-Toque         | 2011              | F. graminearum      | 15-ADON  | 1          |
| 3064 | Cruz Alta            | 2007              | F. graminearum      | 15-ADON  | 1,4,5      |
| 3068 | Ernestina            | 2007              | F. graminearum      | 15-ADON  | 1,2,3,4,5  |
| 3382 | Caseiros             | 2010              | F. austroamericanum | 3-ADON   | 1,2,3,5    |
| 3384 | Barreto              | 2010              | F. austroamericanum | 3-ADON   | 1,2,3,4,5  |
| 3379 | Carazinho            | 2010              | F. cortaderiae      | 3-ADON   | 1,2,3,4,5  |
| 3845 | Lagoa Vermelha       | 2010              | F. cortaderiae      | 3-ADON   | 1,2,3,4    |
| 3841 | Lagoa Vermelha       | 2010              | F. cortaderiae      | 3-ADON   | 1,2,3,4,5  |
| 3831 | Panambi              | 2009              | F. cortaderiae      | NIV      | 1,2,3,4,5  |
| 3836 | Coronel Barros       | 2009              | F. cortaderiae      | NIV      | 1,2,4      |
| 3837 | Coronel Barros       | 2009              | F. cortaderiae      | NIV      | 1,2,3,4,5  |
| 3840 | Mato Castelhano      | 2010              | F. cortaderiae      | NIV      | 1,2,4      |
| 3834 | Barreto              | 2010              | F. cortaderiae      | NIV      | 1,2,4      |
| 3378 | Vacaria              | 2010              | F. cortaderiae      | NIV      | 1,2,3,4,5  |
| 3835 | Ernestina            | 2007              | F. cortaderiae      | NIV      | 1,2,3,4    |
| 3375 | Santa Bárbara do Sul | 2009              | F. meridionale      | NIV      | 1,2,4      |
| 3843 | Ernestina            | 2009              | F. meridionale      | NIV      | 1,2,3,4    |
| 3344 | Coxilha              | 2009              | F. meridionale      | NIV      | 1,2,3,4,5  |
| 3381 | Caseiros             | 2010              | F. meridionale      | NIV      | 1,2,3,4,5  |
| 3839 | Marau                | 2011              | F. meridionale      | NIV      | 1,2,3,4,5  |
| 3377 | Sertão               | 2011              | F. meridionale      | NIV      | 1,2,3,4    |
| 3838 | Nonoai               | 2007              | F. meridionale      | NIV      | 1,2,3,4    |
| 3374 | Nonoai               | 2007              | F. meridionale      | NIV      | 1,2,4      |
